Studies of quantum dots

### 2016-10-06 Manuscript/paper.tex
spEnergiesQD: reorder indices in \sum
Update spEnergiesQD/Manuscript/paper.tex

The harmonic oscillator (HO) is a ubiquitous system in quantum physics. It is
the basis for an extremely diverse set of many-body calculations -- in nuclear
physics \cite{HjorthJensen1995}, quantum chemistry \cite{Helgaker2002}, and
quantum dot calculations of solid state physics \cite{Reimann2002}. In these
calculations, one typically seeks a few of the lowest eigenvalues $E_k$ and
eigenvectors of a many-body Hamiltonian $\hat{H}$ that describes the full
system by solving the Schr\"odinger equation,
\begin{equation*}
\hat{H} \ket{\psi_k} = E_k \ket{\psi_k}, \quad k \in \{1, 2, \ldots, k_{\text{max}}\}.
\end{equation*}
In all cases, the many-body wave function is expanded in a basis of
eigenstates of the one-body HO system.  Necessarily, in order to perform
numerical calculations the basis must be truncated to give an approximation.
In particular, the so-called ``curse of dimensionality'' implies that for a
system with many particles the number of available degrees of freedom per
particle is severely limited by the finiteness of a computer.

@@

The HO eigenfunctions are popular for several reasons. Many quantum systems,
such as the quantum dot model considered here, can be considered perturbed
harmonic oscillators \textit{per se}, thus the true eigenstates should be
perturbations of the HO eigenstates.  Moreover, the HO has many beautiful
properties, such as separability, invariance under orthogonal coordinate
changes, and easily computable eigenfunctions, so that computing matrix
elements of relevant operators becomes relatively simple.  Since HO
eigenfunctions are defined on the \emph{entire} space in which particles live,
a truncation of the spatial domain is unnecessary.  This avoids one of the
main problems with approaches such as finite difference or finite element
methods.\cite{RamMohan2002}

@@

Since Slater determinants are rather cumbersome to write out in full, we shall
instead work in the more convenient formalism of \textit{second quantization}:
\begin{align*}
  \ket{\varnothing_{p_1, \ldots, p_N}} =
  \hat a_{p_1}^\dagger \cdots \hat a_{p_N}^\dagger \ket{\varnothing}
\end{align*}
Here, $\varnothing$ is the vacuum state that is devoid of particles and
$\hat a_p^\dagger$ is a \textit{creation operator} for the single-particle
state $p$.  If the state $p$ is not yet occupied by a particle,
$\hat a_p^\dagger$ creates such a particle.  Otherwise, it collapses the
entire many-body state to $0$.  The Hermitian conjugate of $\hat a_p^\dagger$
is $\hat a_p$, which is an \textit{annihilation operator} for the state $p$
that has the opposite effect.  For fermions, the field operators $\hat a_p$
and $\hat a_p^\dagger$ satisfy the essential anticommutation relations,
\begin{align*}
  &\{\hat a_p, \hat a_q\} = \{\hat a_p^\dagger, \hat a_q^\dagger\} = 0, &
  &\{\hat a_p, \hat a_q^\dagger\} = [p = q].
\end{align*}
where $[P]$ denotes the Iverson bracket, which returns $1$ if the predicate $P$ is true and $0$ otherwise.  Here, $[p = q]$ is simply an alternative way to denote the Kronecker delta $\delta_{p, q}$, but this notation will be used later for other kinds of predicates as well.

@@

The first correction that we will apply to our solution is that of the
Hartree-Fock (HF) method.  Using the variational principle, we compute an
approximate ground state $\Psi_{\text{HF}}$ by minimizing the energy
\begin{align*}
  E_{\Psi_{\text{HF}}} \equiv \langle \Psi_{\text{HF}} | \hat H | \Psi_{\text{HF}} \rangle
\end{align*}
with respect to $\Psi_{\text{HF}}$, subject to the restriction that
$\Psi_{\text{HF}}$ is a single Slater determinant constructed from an unknown
set of $N$ orthonormal single-particle states.  We denote each unknown
single-particle state $\psi_u$ by some label $u$.

To make this computationally feasible, we further assume that the unknown
states $\{\psi_u\}$ are built from a linear combination of known functions,
drawn from a finite set $\{\phi_p\}$.  This reduces the problem from an
abstract mathematical one to a numerical linear algebra problem, with the
caveat that this set of known functions must be chosen carefully to ensure
convergence.

Assuming the interaction is not too strong, we can use the single-particle
states $\{\phi_p\}$ of our noninteracting Hamiltonian as a basis for the
unknown states $\{\psi_u\}$.  The two sets of states are then related by a
linear transformation matrix $\bm C$:
\begin{align*}
  \psi_u \equiv \sum_p \phi_p C_{p, u}
\end{align*}
To ensure the orthonormality of the states, we assume the set of unknown
states $\{\psi_u\}$ has as many elements as the set of $\{\phi_p\}$ and
require the coefficient matrix $\bm C$ to be unitary.  While these conditions
are more strict than necessary, they greatly simplify the calculations and
allow the states $\{\psi_u\}$ to act as inputs for methods beyond HF.  The set
of states $\psi$ now consists of $N$ \textit{occupied} states that participate
in the Slater determinant $\Psi_{\text{HF}}$ and the remaining
\textit{unoccupied} states that do not.

Since the coefficient matrix uniquely determines the unknown states, the
variational problem has been reduced to that of solving of the coefficients
$\bm C$ that minimize the energy $E_{\Psi_{\text{HF}}}$, which is given by
\begin{align}
  E_{\Psi_{\text{HF}}} &= \sum_{u \in \text{occ.}} C_{p, u}^* H_{p, q}^{\text o} C_{q, u}^{}
  \notag \\ &\hphantom{{}=} + \sum_{u, v \in \text{occ.}}
  C_{p, u}^* C_{q, v}^* V_{p, q, r, s}^{} C_{r, u}^{} C_{s, v}^{} \label{eq:hfenergy}
\end{align}

Using the method of Lagrange multiplers, the minimization problem can be
reduced to the solving of a nonlinear equation -- the \textit{Hartree-Fock
  equation}:
\begin{align} \label{eq:hartreefock}
  \bm F \bm C = \bm C \bm \epsilon
\end{align}
where the \textit{Fock matrix} $\bm F$ is defined as
\begin{align} \label{eq:fock}
  F_{p, q} \equiv H_{p, q}^{\text o} + \sum_{r, s \in \text{any}; u \in \text{occ.}}
  C_{r, u}^* V_{p, r, q, s}^{} C_{s, u}^{}
\end{align}
and $\bm \epsilon$ is a vector of Lagrange multipliers with each element
$\epsilon_u$ associated with a particular state $\psi_u$.  [[move this sentence
later to the intro to normal-ordering: While the Fock matrix $\bm F$ may
appear to be an artificial Hamiltonian derived purely from the minimization
problem, it can be considered the normal-ordered operator with respect to a
non-vacuum state, as explained later.]]

The HF equation can be solved by an iterative technique: a solution $\bm C$
can be fed into \eqref{eq:fock} and new solution can then obtained by solving
the eigenvalue problem in \eqref{eq:hartreefock} using the Fock matrix
obtained.  For the initial guess, we use the ground state of our
noninteracting Hamiltonian, thus $\bm C$ is initially the identity matrix.

@@

\subsection{The SRG methods}
\label{subsec:imsrgmethod}

\subsubsection*{Similarity renormalization group methods}

@@

The sequence of transformations is parameterized by a continuous variable $s$
known as the \textit{flow parameter}.  By convention we define to be $0$ at
the beginning of the sequence, and thus the evolving Hamiltonian $\hat H(s)$
at $s = 0$ is set to the original Hamiltonian.  Upon reaching $s$, the
transformed Hamiltonian is given by
\begin{align*}
  \hat H(s) \equiv \hat U(s) \hat H(0) \hat U^\dagger(s)
\end{align*}
where $U(s)$ is a unitary operator that describes the product of all such
transformations since $s = 0$.  Taking the derivative with respect to $s$, we obtain:
\begin{align*}
  \frac{\D}{\D s} H(s) =
  \left(\frac{\D}{\D s} \hat{U}(s)\right) \hat{H}(0) \hat{U}^\dagger(s) +
  \hat{U}(s) \hat{H}(0) \frac{\D}{\D s} \hat{U}^\dagger(s)
\end{align*}
If we define an operator $\eta(s)$ as
\begin{align*}
  \eta(s) \equiv \left(\frac{\D}{\D s} \hat{U}(s)\right) \hat{U}^\dagger(s)
\end{align*}
we find that it is antihermitian as a result of the unitarity of $\hat{U}(s)$:
\begin{align*}
  \eta(s) + \eta^\dagger(s)
  = \frac{\D}{\D s} \left(\hat{U}(s) \hat{U}^\dagger(s)\right)
  = 0
\end{align*}
From this property we can derive a differential equation known as the
\textit{SRG flow equation}:
\begin{align} \label{eq:imsrgode}
  \frac{\D}{\D s} H(s) = [\eta(s), \hat{H}(s)]
\end{align}
The equation allows $\hat{H}(s)$ to be evaluated without explicitly
constructing the full transformation $\hat U(s)$.  The operator $\hat \eta(s)$
is known as the \textit{generator} of the transformation.  When ``integrated''
as a product, the full transformation $\hat U(s)$ is recovered:
\begin{align*}
  \hat U(s')
  = \mathcal T_s \left\{ \mathrm{e}^{\int_0^{s'} \hat{\eta}(s) \D s} \right\}
  \equiv \lim_{\Delta s \to 0} \prod_{s = 0, \Delta s, \ldots}^{s = s'}
  \mathrm{e}^{\hat \eta(s) \Delta s}
\end{align*}
Here $\mathcal T_s$ is a notational device that denotes $s$-ordering: a
reordering of the $\eta(s)$ operator products expanded from the exponential
with respect to $s$ in ascending order \cite[\S 6.1]{reimann2013quantum}.
This is analogous to the usual time-ordering from quantum field theory.

The operator $\hat{\eta}(s)$ can be chosen to suppress certain undesirable
parts of the Hamiltonian matrix, which can be considered the ``off-diagonal''
parts in a loose sense.  The ``off-diagonal'' parts could be elements far away
from the matrix diagonal, in which case the evolution drives the matrix
towards a band-diagonal form.  Or, the ``off-diagonal'' parts could be
elements that couple the ground state from the excited state, in which case
the evolution drives the matrix towards a block-diagonal form that isolates
the ground state.  Or, the ``off-diagonal'' could be literally the elements
that do not lie on the diagonal, in which case the evolution would simply
diagonalize the Hamiltonian.  Through different choices of $\hat{\eta}_s$, the
SRG evolution can be controlled and adapted to the features of a particular
problem.

\subsubsection*{In-medium SRG}

@@

The second-quantized operators in
\eqref{eq:second_quantized_one_body_hamiltonian} and
\eqref{eq:second_quantized_two_body_hamiltonian} are constructed with respect
to the vacuum state.  That is, the operator strings
$\hat{a}_p^\dagger \hat{a}_q$ and
$\hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_s \hat{a}_r$ both have a vacuum
expectation value of zero:
\begin{align*}
&\bra{\varnothing} \hat{a}_p^\dagger \hat{a}_q \ket{\varnothing} = 0 &
&\bra{\varnothing} \hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_s \hat{a}_r \ket{\varnothing} = 0
\end{align*}
They are said to be \emph{normal-ordered} with respect to the vacuum state
$\varnothing$.  It is possible to be normal-ordered with respect to a
non-vacuum state, however.  Consider a Slater determinant $\Phi$ with
$N_{\text F}$ occupied states:
\begin{align*}
  \ket{\Phi} \equiv \ket{\varnothing_{i_1, i_2, \ldots, i_{N_{\text F}}}}
\end{align*}
Here $\{i_1, i_2, \ldots, i_{N_{\text F}}\}$ is a set of $N_{\text F}$ indices that label the occupied states.

One can define a set of \textit{quasiparticle
  field operators} $\hat b$ and $\hat b^\dagger$ that obey the same
anticommutation relations as the $\hat a$ operators.  The field operators are
defined as:
\begin{align*}
  \hat b_p \equiv \begin{cases}
    \hat a_{p} & \text{if $p$ is unoccupied in $\Phi$} \\
    \hat a_{p}^\dagger & \text{if $p$ is occupied in $\Phi$}
  \end{cases}
\end{align*}
The quasiparticle field operators consider $\ket{\Phi}$ to be their ``vacuum state'' (sometimes called ``Fermi vaccuum''), even though it is not a true vacuum state.  Analogously, it is possible to construct any other Slater determinant by applying quasiparticle operators on top of the reference state:
\begin{align*}
  \ket{\Phi_{p_1, \ldots, p_k}} \equiv \hat{b}^\dagger_{p_1} \cdots \hat{b}^\dagger_{p_k} \ket{\Phi}
\end{align*}
In nonrelativistic many-body theory, it is expected that $\{p_1, \ldots, p_k\}$ contain a balanced number of occupied and unoccupied states to conserve the total number of particles.

We can then define normal ordering with respect to the reference state $\Phi$: namely, rearrange an operator string such that every quasiparticle creation operator appears before a quasiparticle annihilation operator by repeatedly exchanging neighboring operators, with each exchange costing a minus sign.  In this form, it is guaranteed that the expectation value of the operator string with respect to
$\Phi$ is zero:
\begin{align*}
&\bra{\Phi} \hat{b}_p^\dagger \hat{b}_q \ket{\Phi} = 0 &
&\bra{\Phi} \hat{b}_p^\dagger \hat{b}_q^\dagger \hat{b}_s \hat{b}_r \ket{\Phi} = 0
\end{align*}
We denote normal ordering of an operator string with respect to $\Phi$ by a
pair of colons:
\begin{align*}
  \normord{\hat{b}_p \hat{b}_q^\dagger} =
  -\hat{b}_q^\dagger \hat{b}_p
\end{align*}
It is not particularly interesting to examine normal ordering of the
quasiparticle field operators, since the result is trivial to compute.  What
is more interesting is the normal ordering of the physical field operators
$\hat a$ and $\hat a^\dagger$ with respect to $\Phi$:
\begin{align*}
  \normord{\hat{a}_p^\dagger \hat{a}_q} &=
  \hat{a}_p^\dagger \hat{a}_q - [p, q \in \text{occ.} \text{ and } p = q]
\end{align*}
This makes the normal ordering notation a useful and concise way of expressing the right-hand side.

Using the normal ordering formalism, we can rewrite the Hamiltonian
$\hat H = \hat H^{\text o} + \hat V$ as
\begin{align}
  \hat{H} = E_\Phi + \sum_{p, q} F_{p, q} \normord{\hat{a}_p^\dagger \hat{a}_q} + \frac{1}{4} \sum_{p, q, r, s} V_{p, q, r, s} \normord{\hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_r \hat{a}_s}
  \label{eq:normordhamiltonian}
\end{align}
where
\begin{align*}
  E_\Phi &= \sum_{i \in \text{occ.}} H^{\text o}_{i, i} + \frac{1}{2} \sum_{i, j \in \text{occ.}} V_{i, j, i, j} \\
  F_{p, q} &= H^{\text o}_{p, q} + \sum_{i \in \text{occ.}} V_{p, i, q, i} \\
\end{align*}
Notice that $E_\Phi$ is in fact the Hartree-Fock energy and $F_{p, q}$ is the Fock matrix.  The representation of $\hat H$ in \eqref{eq:normordhamiltonian} is completely equivalent to that of $\hat H$ in \eqref{eq:vacuumhamiltonian}.  The only difference is that it shuffles the matrix elements so as to put the reference state $\ket{\Phi}$ into the main focus, as opposed to the vacuum state $\ket{\varnothing}$.  Where this makes a crucial difference is when the equations become \emph{truncated}.

Integrating the flow equations (\ref{eq:imsrgode}), we face one of the
major challenges of the SRG method, namely the generation of higher and higher
order interaction terms during the flow.  With each evaluation of the
commutator, the Hamiltonian gains terms of higher order, and these induced
contributions will in subsequent integration steps contribute to terms of
lower order.  In principle, this continues to infinity.

To make the method computationally possible, we must close the IM-SRG flow
equations: we are thus forced to truncate the equations to a certain order. We
choose to truncate both $\hat{H}(s)$ and $\hat{\eta}(s)$ at the two-body
level, an approach which is referred to as IM-SRG(2).  This normal-ordered
two-body approximation appears to be sufficient in many cases and has yielded
excellent results for several nuclei
\cite{PhysRevLett.106.222502,PhysRevLett.109.052501,IMSRG}.

%[[---]]

The commutator in the flow equations \eqref{eq:imsrgode} guarantees that the
IM-SRG wave function $\hat U(s) \ket{\Phi}$ can be expanded in terms of
linked diagrams only \cite{shavitt2009many,ISI:A1981MN73700014}, which
suggests that IM-SRG is size-extensive.  Regarding the quality of the SRG
results, it means that the error introduced by truncating the many-body
expansions scales linearly with the number of particles $N$.

With this truncation, the generator $\hat{\eta}$ can be written as
\begin{align*}
\hat{\eta} = \sum_{p, q} \eta_{p, q}^{[1]} \normord{\hat a_p^\dagger \hat a_q} +
\frac{1}{4} \sum_{p, q, r, s}\eta_{p, q, r, s}^{[2]} \normord{\hat a_p^\dagger \hat a_q^\dagger \hat a_s \hat a_r}
\end{align*}
where $\eta_{p, q}^{[1]}$ and $ \eta_{p, q, r, s}^{[2]}$ are the one- and two-body elements respectively.  For consistency, we define $H^{[0]} \equiv E_\Phi$, $H^{[1]}_{p, q} \equiv F_{p, q}$, and $H^{[2]}_{p, q, r, s} \equiv V_{p, q, r, s}$ as well.

>>>>

The harmonic oscillator (HO) is a ubiquitous system in quantum physics. It is the basis for an extremely diverse set of many-body calculations -- in nuclear physics \cite{HJORTHJENSEN1995125}, quantum chemistry \cite{Helgaker2002} [[missing ref for Helgaker 2002]], and quantum dot calculations of solid state physics \cite{reimann2002}. In these calculations, one typically seeks a few of the lowest eigenvalues $E_k$ and eigenvectors of a many-body Hamiltonian $\hat{H}$ that describes the full system by solving the Schr\"odinger equation,
\begin{equation*}
\hat{H} \ket{\psi_k} = E_k \ket{\psi_k}, \quad k \in \{1, 2, \ldots, k_{\text{max}}\}.
\end{equation*}
In all cases, the many-body wave function is expanded in a basis of
eigenstates of the one-body HO system.  Necessarily, in order to perform
numerical calculations the basis must be truncated to give an approximation.
In particular, the so-called ``curse of dimensionality'' implies that for a
system with many particles the number of available degrees of freedom per
particle is severely limited by the finiteness of a computer.

@@

The HO eigenfunctions are popular for several reasons. Many quantum systems, such as the quantum dot model considered here, can be considered perturbed harmonic oscillators \textit{per se}, thus the true eigenstates should be perturbations of the HO eigenstates.  Moreover, the HO has many beautiful properties, such as separability, invariance under orthogonal coordinate changes, and easily computable eigenfunctions, so that computing matrix elements of relevant operators becomes relatively simple.  Since HO eigenfunctions are defined on the \emph{entire} space in which particles live, a truncation of the spatial domain is unnecessary.  This avoids one of the main problems with approaches such as finite difference or finite element methods.\cite{RamMohan2002} [[Unknown citation]]

@@

Since Slater determinants are rather cumbersome to write out in full, we shall
instead work in the more convenient formalism of \textit{second quantization}:
\begin{align*}
  \ket{\varnothing_{p_1, \ldots, p_N}} =
  \hat a_{p_1}^\dagger \cdots \hat a_{p_N}^\dagger \ket{\varnothing}
\end{align*}
Here, $\varnothing$ is the vacuum state that is devoid of particles and
$\hat a_p^\dagger$ is a \textit{creation operator} for the single-particle
state $p$.  If the state $p$ is not yet occupied by a particle,
$\hat a_p^\dagger$ creates such a particle.  Otherwise, it collapses the
entire many-body state to $0$.  The Hermitian conjugate of $\hat a_p^\dagger$
is $\hat a_p$, which is an \textit{annihilation operator} for the state $p$
that has the opposite effect.  For fermions, the field operators $\hat a_p$
and $\hat a_p^\dagger$ satisfy the essential anticommutation relations,
\begin{align*}
  &\{\hat a_p, \hat a_q\} = \{\hat a_p^\dagger, \hat a_q^\dagger\} = 0 &
  &\{\hat a_p, \hat a_q^\dagger\} = \delta_{p, q}
\end{align*}
where $\delta_{p, q}$ denotes the Kronecker delta.

@@

The first correction that we will apply to our solution is that of the
Hartree-Fock (HF) method.  Using the variational principle, we compute an
approximate ground state $\Phi$ by minimizing the energy expectation value
\begin{align*}
  E_{\Phi} \equiv \langle \Phi | \hat H | \Phi \rangle
\end{align*}
with respect to $\Phi$, subject to the restriction that
$\Phi$ is a single Slater determinant constructed from an unknown
set of $N$ orthonormal single-particle states.  We denote each unknown
single-particle state $\phi_{q}'$ by some label $q$.

To make this computationally feasible, we further assume that the unknown
states $\{\phi_q'\}$ are built from a linear combination of known functions,
drawn from a finite set $\{\phi_p\}$.  This reduces the problem from an
abstract mathematical one to a numerical linear algebra problem, with the
caveat that this set of known functions must be chosen carefully to ensure
convergence.

Assuming the interaction is not too strong, we can use the single-particle
states $\{\phi_p\}$ of our noninteracting Hamiltonian as a basis for the
unknown states $\{\phi'_u\}$.  The two sets of states are then related by a
linear transformation matrix $\bm C$:
\begin{align*}
  \phi_q' \equiv \sum_p \phi_p C_{p, q}
\end{align*}
To ensure the orthonormality of the states, we assume the set of unknown states $\{\phi_q'\}$ has as many elements as the set of $\{\phi_p\}$ and require the coefficient matrix $\bm C$ to be unitary.  While these conditions are more strict than necessary, they greatly simplify the calculations and allow the states $\{\phi_q'\}$ to act as inputs for methods beyond HF.  The set of states $\{\phi_q'\}$ now consists of $N$ \textit{occupied} states that participate in the Slater determinant $\Phi$ and the remaining \textit{unoccupied} states that do not.

Since the coefficient matrix uniquely determines the unknown states, the variational problem has been reduced to that of solving of the coefficients $\bm C$ that minimize the energy $E_{\Phi}$, which is given by
\begin{align}
  E_{\Phi} &= \sum_{j} n_j H^{\text{o} \prime}_{j, j} + \sum_{j, k} n_j n_k V'_{j, k, j, k} \label{eq:hfenergy}
\end{align}
where the transformed operators are defined as
\begin{align}
  H^{\text{o} \prime}_{p, q} &\equiv \sum_{r, s} C_{r, p}^* H_{r, s}^{\text o} C_{s, q}^{} \label{eq:hftransform1} \\
  V'_{p, q, r, s} &\equiv \sum_{t, u, v, w} C_{t, p}^* C_{u, q}^* V_{t, u, v, w}^{} C_{v, r}^{} C_{w, s}^{} \label{eq:hftransform2}
\end{align}
and the occupancy number $n_p$ is defined as
\begin{align}
  n_p \equiv \begin{cases}
    1 & \text{if $p$ is an occupied state in $\Phi$} \\
    0 & \text{if $p$ is an unoccupied state in $\Phi$}
  \end{cases}
\end{align}
and further define $\bar n_p \equiv 1 - n_p$ as the complement.

With the method of Lagrange multipliers, the minimization problem can be reduced to the solving of a nonlinear equation -- the \textit{Hartree-Fock equation}:
\begin{align} \label{eq:hartreefock}
  \bm F \bm C = \bm C \bm \epsilon
\end{align}
where the \textit{Fock matrix} $\bm F$ is defined as
\begin{align} \label{eq:fock}
  F_{p, q} \equiv H_{p, q}^{\text o} + \sum_{r, s, j} n_j C_{r, j}^* V_{p, r, q, s}^{} C_{s, j}^{}
\end{align}
and $\bm \epsilon$ is a vector of Lagrange multipliers with each element $\epsilon_q$ associated with a particular state $\phi'_q$.

The HF equation can be solved by an iterative technique: a solution $\bm C$ can be fed into \eqref{eq:fock} and new solution can then obtained by solving the eigenvalue problem in \eqref{eq:hartreefock} using the Fock matrix obtained.  For the initial guess, we use the ground state of our noninteracting Hamiltonian, thus $\bm C$ is initially the identity matrix.  After dozens of iterations, perhaps with the aid of a convergence acceleration technique (such as DIIS [[cite??] or Broyden's method [[cite?]]), one would usually reach a state of self-consistency where the solution converges to a fixed value.  This is the Hartree-Fock solution.

@@

To make use of the HF solution as the reference state for future calculations, we transform the operators via \eqref{eq:hftransform1} and \eqref{eq:hftransform2}.  Therefore, the post-HF methods described in the next few sections generally use the transformed operators $\hat H^{\text{o} \prime}$ and $\hat V'$ as inputs.  However, for generality, we omit the prime symbols as the methods do not necessitate a state optimized by HF.

\subsection{The IM-SRG method}
\label{subsec:imsrgmethod}

\subsubsection{Similarity renormalization group methods}
\label{subsubsec:srgmethods}

@@

The sequence of transformations is parameterized by a continuous variable $s$
known as the \textit{flow parameter}.  By convention we define to be $0$ at
the beginning of the sequence, and thus the evolving Hamiltonian $\hat H(s)$
at $s = 0$ is set to the original Hamiltonian.  Upon reaching $s$, the
transformed Hamiltonian is given by
\begin{align*}
  \hat H(s) \equiv \hat U(s) \hat H(0) \hat U^\dagger(s)
\end{align*}
where $U(s)$ is a unitary operator that describes the product of all such
transformations since $s = 0$.  Taking the derivative with respect to $s$, we obtain:
\begin{align*}
  \frac{\D}{\D s} H(s) =
  \left(\frac{\D}{\D s} \hat{U}(s)\right) \hat{H}(0) \hat{U}^\dagger(s) +
  \hat{U}(s) \hat{H}(0) \frac{\D}{\D s} \hat{U}^\dagger(s)
\end{align*}
If we define an operator $\eta(s)$ as
\begin{align*}
  \eta(s) \equiv \left(\frac{\D}{\D s} \hat{U}(s)\right) \hat{U}^\dagger(s)
\end{align*}
we find that it is antihermitian as a result of the unitarity of $\hat{U}(s)$:
\begin{align*}
  \eta(s) + \eta^\dagger(s)
  = \frac{\D}{\D s} \left(\hat{U}(s) \hat{U}^\dagger(s)\right)
  = 0
\end{align*}
From this property we can derive a differential equation known as the
\textit{SRG flow equation}:
\begin{align} \label{eq:imsrgode}
  \frac{\D}{\D s} H(s) = [\eta(s), \hat{H}(s)]
\end{align}
The equation allows $\hat{H}(s)$ to be evaluated without explicitly
constructing the full transformation $\hat U(s)$.  The operator $\hat \eta(s)$
is known as the \textit{generator} of the transformation.  When ``integrated''
as a product, the full transformation $\hat U(s)$ is recovered:
\begin{align*}
  \hat U(s')
  = \mathcal T_s \left\{ \mathrm{e}^{\int_0^{s'} \hat{\eta}(s) \D s} \right\}
  \equiv \lim_{\Delta s \to 0} \prod_{s = 0, \Delta s, \ldots}^{s = s'}
  \mathrm{e}^{\hat \eta(s) \Delta s}
\end{align*}
Here $\mathcal T_s$ is a notational device that denotes $s$-ordering: a reordering of the $\hat{\eta}(s)$ operator products expanded from the exponential with respect to $s$ in ascending order \cite[\S 6.1]{reimann2013quantum}.  This is analogous to the usual time-ordering from quantum field theory.

The operator $\hat{\eta}$ can be chosen to suppress certain undesirable parts of the Hamiltonian matrix, which can be considered the ``off-diagonal'' parts in a loose sense.  The ``off-diagonal'' parts could be elements far away from the matrix diagonal, in which case the evolution drives the matrix towards a band-diagonal form.  Or, the ``off-diagonal'' parts could be elements that couple the ground state from the excited state, in which case the evolution drives the matrix towards a block-diagonal form that isolates the ground state.  Or, the ``off-diagonal'' could be literally the elements that do not lie on the diagonal, in which case the evolution would simply diagonalize the Hamiltonian.  Through different choices of $\hat{\eta}$, the SRG evolution can be controlled and adapted to the features of a particular problem.

\subsubsection{Evolving the flow equation in medium}

@@

The second-quantized operators in
\eqref{eq:second_quantized_one_body_hamiltonian} and
\eqref{eq:second_quantized_two_body_hamiltonian} are constructed with respect
to the vacuum state.  That is, the operator strings
$\hat{a}_p^\dagger \hat{a}_q$ and
$\hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_s \hat{a}_r$ both have a vacuum
expectation value of zero:
\begin{align*}
&\bra{\varnothing} \hat{a}_p^\dagger \hat{a}_q \ket{\varnothing} = 0 &
&\bra{\varnothing} \hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_s \hat{a}_r \ket{\varnothing} = 0
\end{align*}
They are said to be \emph{normal-ordered} with respect to the vacuum state
$\varnothing$.  It is possible to be normal-ordered with respect to a
non-vacuum state, however.  Consider a Slater determinant $\Phi$ with
$N_{\text F}$ occupied states:
\begin{align*}
  \ket{\Phi} \equiv \ket{\varnothing_{i_1, i_2, \ldots, i_{N_{\text F}}}}
\end{align*}
where $\{i_1, i_2, \ldots, i_{N_{\text F}}\}$ is a set of $N_{\text F}$ indices that label the occupied states.  One can define a set of \textit{quasiparticle field operators} $\hat b$ and $\hat b^\dagger$ that obey the same anticommutation relations as the $\hat a$ operators:
\begin{align*}
  \hat b_p \equiv \bar n_p \hat a_{p} + n_p \hat a_{p}^\dagger
\end{align*}
The quasiparticle field operators treat $\ket{\Phi}$ as their ``vacuum state'' (sometimes called ``Fermi vaccuum'' in an abstract sense), despite not being the physical vacuum state.  As before, it is possible to construct arbitrary Slater determinants by applying quasiparticle operators to the reference state:
\begin{align*}
  \ket{\Phi_{p_1, \ldots, p_k}} \equiv \hat{b}^\dagger_{p_1} \cdots \hat{b}^\dagger_{p_k} \ket{\Phi}
\end{align*}
In nonrelativistic many-body theory, it is expected that $\{p_1, \ldots, p_k\}$ contain a balanced number of occupied and unoccupied states to conserve the total number of particles; otherwise, the Slater determinant would represent a different system.

We may then define \emph{normal ordering with respect to the reference state} $\Phi$, a notational device that rearranges a string of operators such that every quasiparticle creation operator appears before a quasiparticle annihilation operator by repeatedly exchanging neighboring operators, with each exchange incurring a change in sign.  A simple example would be the normal ordering of a pair of field operators such as:
\begin{align*}
  \normord{\hat{b}_p \hat{b}_q^\dagger} =
  -\hat{b}_q^\dagger \hat{b}_p
\end{align*}
Here, we denote normal ordering of an operator string with respect to $\Phi$ by a pair of colons.  The expectation value of a normal ordered string of operators with respect to $\Phi$ is guaranteed to vanish:
\begin{align*}
&\bra{\Phi} \hat{b}_p^\dagger \hat{b}_q \ket{\Phi} = 0 &
&\bra{\Phi} \hat{b}_p^\dagger \hat{b}_q^\dagger \hat{b}_s \hat{b}_r \ket{\Phi} = 0
\end{align*}
The normal ordering of the quasiparticle field operators are straightforward to compute, but what is more interesting is the normal ordering of the physical field operators $\hat a$ and $\hat a^\dagger$ with respect to $\Phi$:
\begin{align*}
  \normord{\hat{a}_p^\dagger \hat{a}_q} &=
  \hat{a}_p^\dagger \hat{a}_q - n_p \delta_{p, q}
\end{align*}
The normal ordering notation thus provides a concise way of expressing the right-hand side.  In this new notation, we can rewrite the Hamiltonian $\hat H = \hat H^{\text o} + \hat V$ in its \emph{normal-ordered representation}:
\begin{align}
  \hat{H} = H^{[0]} + \sum_{p, q} H^{[1]}_{p, q} \normord{\hat{a}_p^\dagger \hat{a}_q} + \frac{1}{4} \sum_{p, q, r, s} H^{[2]}_{p, q, r, s} \normord{\hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_r \hat{a}_s}
  \label{eq:normordhamiltonian}
\end{align}
where
\begin{align*}
  H^{[0]} &\equiv \sum_{i} n_i H^{\text o}_{i, i} + \frac{1}{2} \sum_{i, j} n_i n_j V_{i, j, i, j} \\
  H^{[1]}_{p, q} &\equiv H^{\text o}_{p, q} + \sum_{i} n_i V_{p, i, q, i} \\
  H^{[2]}_{p, q, r, s} &\equiv V_{p, q, r, s} \\
\end{align*}
If $\Phi$ is a state optimized by the HF method, then $H^{[0]}$ is simply the HF energy $E_\Phi$ and $H^{[1]}_{p, q}$ is the Fock matrix $F_{p, q}$.  The normal-ordered representation of $\hat H$ in \eqref{eq:normordhamiltonian} is completely equivalent to that of $\hat H$ in \eqref{eq:vacuumhamiltonian}.  The only difference is that it rearranges the matrix elements so as to emphasize the reference state $\Phi$ rather than vacuum state $\ket{\varnothing}$.  This makes a critical difference when the expressions containing 3-body or higher terms are \emph{truncated} to only 2-body.  If the reference state $\Phi$ is good approximation of the true solution $\Psi$, then the importance of the higher-body parts of the Hamiltonian are greatly reduced.

Higher-body operators arise from integrating the flow equations (\ref{eq:imsrgode}), which is one of the major challenges of the SRG method.  With each evaluation of the commutator, the Hamiltonian gains terms of higher order, and these induced contributions will in subsequent integration steps feed back into terms of lower order.  Thus, the higher-body contributions are not irrelevant to the final solution even if only the ground state energy is of interest.

Computationally, higher-body terms are difficult to handle: the amount of memory required to store an $r$-body operator is grows exponentially with $r$.  Moreover, the flow equations are capable of generating an infinite number of higher-body terms as the Hamiltonian evolves.  Thus, to make the method tractable, we must close the IM-SRG flow equations, truncating the equations to a finite order.  In this paper, we truncate both $\hat{H}$ and $\hat{\eta}$ at the two-body level, an approach known as IM-SRG(2).  This normal-ordered two-body approximation appears to be sufficient in many cases and has yielded excellent results for several nuclei \cite{PhysRevLett.106.222502,PhysRevLett.109.052501,IMSRG}.

With the truncation, the generator $\hat{\eta}$ can be written as a generic antihermitian 2-body operator:
\begin{align*}
\hat{\eta} = \sum_{p, q} \eta_{p, q}^{[1]} \normord{\hat a_p^\dagger \hat a_q} +
\frac{1}{4} \sum_{p, q, r, s}\eta_{p, q, r, s}^{[2]} \normord{\hat a_p^\dagger \hat a_q^\dagger \hat a_s \hat a_r}
\end{align*}
where $\eta_{p, q}^{[1]}$ and $ \eta_{p, q, r, s}^{[2]}$ respectively are its one- and two-body matrix elements normal ordered with respect to $\Phi$.
